% =============================================================================
% Cover Letter for the NL2Semantic2SQL submission to IJGIS.
% Anonymised for double-anonymised review: no authors, no affiliations,
% no signature line. Corresponding author is identified via the editorial
% submission system.
% =============================================================================
\documentclass[11pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{url}
\usepackage[hidelinks]{hyperref}

\address{Anonymous corresponding author\\ \textit{(Identity provided to editors through the submission system; withheld for double-anonymised review)}}

\begin{document}

\begin{letter}{The Editor-in-Chief \\ \emph{International Journal of Geographical Information Science}}

\opening{Dear Editor,}

We submit for your consideration the manuscript ``\textbf{Semantic Grounding and Safe Execution for PostGIS Natural-Language-to-SQL},'' a regular research article for \emph{IJGIS}. The present submission is anonymised for double-anonymised peer review; author identity is provided through the editorial system.

\medskip
\noindent\textbf{Fit to IJGIS.} The manuscript addresses a GIScience question: how to ground natural-language queries so that a large language model can produce executable PostGIS SQL that respects Simple Feature semantics, CRS transformations, the geometry/geography distinction, KNN operators, and safety constraints. Its primary contribution is on the spatial-database side; BIRD is used as a second evaluation track for the warehouse path of the same bilingual framework. The grounding formalisms we encode trace to the $4$- and $9$-intersection topological relations (Egenhofer and Franzosa, 1991; Clementini et al., 1993), the OGC Simple Feature Access standard (\emph{not} a generic LLM benchmark), and the executable semantics of PostGIS.

\medskip
\noindent\textbf{What the paper claims, and at what confidence level.}

\begin{enumerate}
\item \textbf{Primary GIS results (calibrated with $N=3$ Spatial sampling; Robustness is the strongest single claim).} On the 40-question Robustness suite over live Chongqing spatial data, the full pipeline reaches $39/40 = 0.975$ vs.\ a direct-LLM baseline of $18/40 = 0.450$ (paired McNemar $b=0$, $c=21$, $p<10^{-4}$); OOM Prevention bounded-answer compliance improves from $1/8$ to $7/8$ after an \texttt{EXPLAIN}-based pre-check and a revised bounded-output agent prompt. On the 85-question Spatial set we report three independent Full-mode samples (EX = 0.706, 0.612, 0.671; mean$\pm$SD = $0.663\pm 0.048$; majority-vote = 0.659, paired McNemar $p=0.052$). Two of three samples individually reach significance ($p=0.003$, $p=0.029$); one does not ($p=0.230$). The $N=3$ protocol and honest reporting of sample variance replaces the earlier single-draw figure.

\item \textbf{Robustness paired, not Full-only.} The 40-question Robustness suite is reported as a paired comparison against baseline (the earlier Full-only framing has been replaced). Full reaches 39/40 = 0.975 vs.\ baseline 18/40 = 0.450 (paired McNemar $b=0$, $c=21$, $p<10^{-4}$). Per-category Full accuracy: Security Rejection 11/11, Anti-Illusion 9/9, Schema Enforcement 3/3, Schema Hallucination 8/8, Data Tampering 1/1, OOM Prevention 7/8. The one remaining OOM failure is a specific Chinese phrasing where the agent still defers to semantic-layer ambiguity; this is discussed in the limitations section rather than claimed as solved.

\item \textbf{BIRD as the second evaluation track of the same bilingual framework, not as a second primary domain.} We report BIRD mini\_dev in two explicitly separated parts: a 108-question \emph{design set} used to derive grounding rules by error attribution on round-1 failures, and a disjoint 150-question \emph{held-out set} sampled from the remaining BIRD mini\_dev PostgreSQL items. The design set improves by $+0.093$ with $p=0.0213$, which we report as a development-set tuning effect rather than a generalisation claim. On the independent held-out set, the same rules produce only $+0.033$ with 95\% CI $[-0.03, 0.09]$ and $p=0.3833$; we do not claim a BIRD-wide improvement. This is a current-state snapshot of the warehouse path and is what scopes our next round of warehouse-side grounding work; the calibrated reporting is a deliberate response to external peer review on the earlier version.

\item \textbf{Cross-lingual gap is not a translation artefact.} A 50-question Chinese BIRD probe is reported alongside a bilingual human re-audit of all 50 LLM translations (18~ok, 32~fix, 0~drop) and a paired re-run of the Full pipeline on the reviewed set. Pre-review EX $0.560$ vs.\ post-review EX $0.540$ (paired McNemar discordant pairs $b=2$, $c=1$, two-sided exact $p=1.00$) rules out translation artefacts as the cause of the English$\to$Chinese degradation. The remaining cross-lingual gap therefore reflects genuine cross-lingual reasoning and schema-linking difficulty, scoping the next cross-lingual engineering round to bilingual grounding rather than translation improvement.
\end{enumerate}

\medskip
\noindent\textbf{Changes relative to any earlier version.} An earlier version of this work received external review. The present submission responds substantively to that review and to two additional reviews received on a prior submission, delivering 6 P0 reviewer fixes and an extensive methodological deepening: (i)~the title is narrowed to focus on PostGIS natural-language-to-SQL; (ii)~the semantic layer is formalised as a metadata graph $G=(V,E)$ with a rule-to-OGC-SFA alignment table, addressing the theoretical-depth feedback from an earlier peer-review report; (iii)~an agent-loop-native ablation and a schema-only middle baseline are added to isolate the contribution of semantic grounding from generic agent scaffolding; (iv)~an EXPLAIN-based OOM pre-check is implemented and the Robustness OOM bounded-answer compliance is updated accordingly (1/8 $\rightarrow$ 7/8); (v)~Spatial EX is re-measured with $N=3$ independent samples at temperature $=0.0$ and reported as pooled mean$\pm$SD with majority-vote aggregation, calibrating the significance claim; (vi)~a bilingual human re-audit of all 50 Chinese BIRD translations and a paired re-run of the full pipeline on the reviewed set is now landed in §4.9 (paired $p=1.00$), ruling out translation-artefact as the cause of the cross-lingual gap; (vii)~a benchmark representativeness profile (predicate family and PostGIS function distribution) is added as Table~1b; (viii)~a direct comparison table to recent NL-to-spatial-database systems (NALSpatial, Monkuu, GeoSQL-Eval, GeoCogent, GeoAgent) is added; (ix)~the execution-accuracy evaluator is documented with numeric tolerance, NULL handling, and LIMIT semantics; (x)~all previously cited references with metadata issues are corrected (notably Monkuu, PostGIS manual, OGC SFA/ISO). A separate response-to-reviewers document accompanies this submission.

\medskip
\noindent\textbf{Reproducibility and anonymisation.} The 125-question Chongqing PostGIS benchmark with gold SQL, the semantic-layer YAML, the intent-rule set, the prompt templates, the SQL postprocessor, the DIN-SQL runner adapted for PostgreSQL/PostGIS, the six-way single-pass diagnostic runner, and the per-question predictions backing every table in the paper are released in an anonymised companion repository. An anonymised reviewer-accessible link is provided through the editorial system. All runs use a single LLM (Gemini~2.5~Flash) at temperature $=0.0$; the paper is explicit about this and lists it as a limitation.

\medskip
\noindent\textbf{Originality, conflicts of interest, and double-anonymised compliance.} The manuscript is original work and has not been published or submitted elsewhere. No portion has appeared in any conference or workshop. The author(s) declare no competing interests. In accordance with IJGIS's double-anonymised policy, all author names, affiliations, acknowledgements, and self-citations that would identify the authors have been removed; commit hashes and institutional URLs have been replaced with anonymised equivalents.

We thank you for considering this submission and look forward to the reviewers' feedback.

\closing{Sincerely,}

\end{letter}

\end{document}
